Temporal ordering and registration of images in studies of developmental dynamics

### 2014-07-10 image_analysis_paper/SI_text.tex
Added live imaging movies

\bibliographystyle{pnas2}
\bibliography{background_reading/references,../../references/references}

@@

\begin{figure}
\includegraphics[width=8.4cm]{ap_dv}
\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green). The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right. The arrow indicates the position(s) where the cross-section of an embryo is imaged. (Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo. The dorsal (D) side is up and the ventral (V) side is down. For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image). During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{...}. In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image). }
\label{fig:ap_dv}
\end{figure}

\begin{figure}
\includegraphics[width=8.4cm, trim=0cm 5.6cm 0cm 0cm, clip]{SI_fig6}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development. The membrane length is indicated by the red bars. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows.}
\label{fig:membrane_compare}
\end{figure}

\begin{figure}
\includegraphics[height=9cm]{raw_data3}
\caption{Images of mutant and wild type {\it Drosophila} embryos during cellularization, used for the analysis in \fig~5. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half are of mutant embryos.}
\end{figure}

\begin{figure}
\includegraphics[width=8.4cm]{SI_fig3}
\caption{Synthetic data set used to illustrate the data processing algorithms. {\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). {\it (B)} Intensity corresponding to the profile in {\it A}. {\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}. Each row in the array corresponds to a single profile. {\it (D)} The profiles in {\it C}, now registered using angular synchronization. {\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps. {\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
\label{fig:1d_demo}
%\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
\end{figure}

\begin{figure*}
\begin{minipage}{0.22\textwidth}
\centering
%\includegraphics[width=\textwidth]{sphere_1}\\
\includegraphics[width=\textwidth]{sphere2_1}\\
\sffamily{A}
\end{minipage}
%
\hfill
%
\begin{minipage}{0.22\textwidth}
\centering
%\includegraphics[width=\textwidth]{sphere_2}\\
\includegraphics[width=\textwidth]{sphere2_2}\\
\sffamily{B}
\end{minipage}
\begin{minipage}{0.22\textwidth}
\centering
%\includegraphics[width=\textwidth]{sphere_3}\\
\includegraphics[width=\textwidth]{sphere2_3}\\
\sffamily{C}
\end{minipage}
\begin{minipage}{0.22\textwidth}
\centering
%\includegraphics[width=\textwidth]{sphere_4}\\
\includegraphics[width=\textwidth]{sphere2_4}\\
\sffamily{D}
\end{minipage}
\caption{Illustration of the correspondence between three-dimensional rotations to translations and rotations in two dimensions. {\it (A)} The original image (shown in red), projected onto a portion of a sphere. {\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. {\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.}
\label{fig:SO3_picture}
\end{figure*}

\begin{figure*}
%\raisebox{4.5cm}{\sffamily{A}}
%\includegraphics[width=5cm]{data1_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data1_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{A}};
%	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
	\draw[<->, ultra thick, cyan] (-1.4cm, -1.85cm) -- (1.7cm, -1.85cm);
	\draw[->, ultra thick, red] (1.97cm, -0.5cm) -- (1.97cm, -1cm);
\end{tikzpicture}
%\raisebox{4.5cm}{\sffamily{B}}
%\includegraphics[width=5cm]{data2_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data2_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{B}};
	\draw[<->, ultra thick, cyan] (-1.2cm, -1.85cm) -- (1.5cm, -1.85cm);
	\draw[->, ultra thick, red] (1.97, -0.85cm) -- (1.97cm, -1.34cm);
\end{tikzpicture}
%\raisebox{4.5cm}{\sffamily{C}}
%\includegraphics[width=5cm]{data3_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data3_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{C}};
	\draw[<->, ultra thick, cyan] (-0.25cm, -1.85cm) -- (0.35cm, -1.85cm);
	\draw[->, ultra thick, red] (1.97, -1.1cm) -- (1.97cm, -1.6cm);
	\draw[->, ultra thick, red] (1.52, -1.1cm) -- (1.52cm, -1.6cm);
	\draw[->, ultra thick, red] (0.64, -1.1cm) -- (0.64cm, -1.6cm);
\end{tikzpicture}
\caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets.
We expect to see an approximately symmetric semicircle distribution centered at 0, which corresponds to noise, and positive eigenvalues separated from the semicircle which correspond to meaningful coordinates.
Each coordinate yields $d=3$ eigenvalues of (approximately) the same magnitude, and the top $d=3$ eigenvalues ($\lambda=1$) correspond to a trivial embedding coordinate.
(A) Histogram of eigenvalues from images during cellularization (\fig~3). As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics; then there is an appreciable gap before the third group of eigenvalues.
(B) Histogram of eigenvalues from images during gastrulation (\fig~4). We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics.
(C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are removed from this distribution (indicated by the red arrows). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $6$ eigenvalues correspond to the first two VDM components which are now necessary to parametrize the developmental dynamics of the two genotypes.}
%there is a spectral gap after the first (non-trivial) group of eigenvalues.
%\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets.
%%
%As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations.
%%
%(A) Spectrum from images during cellularization (\fig~3). As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
%%
%(B) Spectrum from images during gastrulation (\fig~4). Again,
%there is a spectral gap after the first (non-trivial) group of eigenvalues.
%%
%(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
%We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data. }
\label{fig:eval_spectra}
\end{figure*}

\begin{figure}
\includegraphics[width=8.7cm]{PCA_ordered}
\caption{Images of embryos during cellularization (from \fig~3), registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient. The recovered dynamics are qualitatively consistent with those recovered by vector diffusion maps. }
\label{fig:PCA_data1}
\end{figure}

\begin{figure*}
\raisebox{5.2cm}{\sffamily{A}}
\includegraphics[width=16.8cm]{data2_PCA_ordered}

\raisebox{3.7cm}{\sffamily{B}}
\includegraphics[width=8.4cm, angle=90]{data2_PCA_variance}
\raisebox{3.7cm}{\sffamily{C}}
\includegraphics[width=8.4cm]{data2_PCA_proj}

@@

%
%\begin{figure*}[t]
%{\bf Item C in the figure below is our main observation.}
%
%\raisebox{7.6cm}{\sffamily{A}}
%\includegraphics[width=8cm]{raw_data1}
%\hfill
%\raisebox{7.6cm}{\sffamily{B}}
%\includegraphics[width=8cm]{FBsPCA_evals}
%
%\raisebox{5.7cm}{\sffamily{C}}
%\includegraphics[width=16cm, trim=2cm 0cm 2cm 0cm, clip]{FBsPCA_dmaps}
%\caption{I computed the FBsPCA basis for the set of images (separately for the red and green signals).
%%
%I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
%%
%The results are shown above.
%%
%However, (mainly, we suspect) because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.
%%
%(A) Images used for FBsPCA. (B) Eigenvalue spectrum from diffusion maps, computed using bispectrum coefficients. There is no visible spectral gap after the first component, as would be expected for this particular data set. (C, top) First 4 diffusion maps variables, computed from the bispectrum coefficients, plotted against the membrane thickness. Note there is no appreciable correlation: {\bf each diffusion maps variable simply selects one image from the data set as ``abnormal''.} (C, bottom) The corresponding images that each diffusion maps mode ``selects.'' In general, these seem to be the late images with ventral furrows beginning to form. }
%\end{figure*}

>>>>

\begin{thebibliography}{10}

\bibitem{singer2011angular}
Singer A (2011) Angular synchronization by eigenvectors and semidefinite
  programming. \textit{Appl Comput Harmon Anal}
  30(1):20--36.

\bibitem{ahuja2007template}
Ahuja S, Kevrekidis I~G, Rowley C~W (2007) Template-based stabilization of
  relative equilibria in systems with continuous symmetry. \textit{Journal of
  Nonlinear Science} 17(2):109--143.

\bibitem{goemans1995improved}
Goemans M~X, Williamson D~P (1995) Improved approximation algorithms for maximum cut and satisfiability problems using semidefinite programming.
\textit{Journal of the ACM} 42(6):1115--1145.

\bibitem{bandeira2013cheeger}
Bandeira A~S, Singer A, Spielman D~A (2013) A Cheeger inequality for the graph connection Laplacian. \textit{SIAM Journal on Matrix Analysis and Applications} 34(4):1611--1630.

\bibitem{coifman2005geometric}
Coifman R~R, Lafon S, Lee A~B, Maggioni M, Nadler B, \textit{et~al.} (2005)
  Geometric diffusions as a tool for harmonic analysis and structure definition
  of data: Diffusion maps. \textit{Proc Natl Acad Sci USA} 102(21):7426--7431.

\bibitem{coifman2008graph}
Coifman R~R, Shkolnisky Y, Sigworth F~J, Singer A (2008) Graph {L}aplacian
  tomography from unknown random projections. \textit{IEEE Trans Image Process} 17(10):1891--1899.

\bibitem{rohrdanz2011determination}
Rohrdanz M~A, Zheng W, Maggioni M, Clementi C (2011) Determination of reaction
  coordinates via locally scaled diffusion map. \textit{J Chem Phys} 134:124116.

\bibitem{Belkin2003}
Belkin M, Niyogi P (2003) Laplacian eigenmaps for dimensionality reduction and
  data representation. \textit{Neural Comput} 15:1373--1396.

\bibitem{singer2012vector}
Singer A, Wu H~T (2012) Vector diffusion maps and the connection laplacian.
  \textit{Commun Pure Appl Math} 65(8):1067--1144.

\bibitem{singer2013spectral}
Singer A, Wu H~t (2013) Spectral convergence of the connection {L}aplacian from
  random samples. \textit{arXiv preprint arXiv:13061587} .

\bibitem{wigner1955characteristic}
Wigner E~P (1955) Characteristic vectors of bordered matrices with infinite
  dimensions. \textit{The Annals of Mathematics} 62(3):548--564.

\bibitem{wigner1958distribution}
Wigner E~P (1958) On the distribution of the roots of certain symmetric
  matrices. \textit{The Annals of Mathematics} 67(2):325--327.

\bibitem{ferguson2010systematic}
Ferguson A~L, Panagiotopoulos A~Z, Debenedetti P~G, Kevrekidis I~G (2010)
  Systematic determination of order parameters for chain dynamics using
  diffusion maps. \textit{Proc Natl Acad Sci USA}
  107(31):13597--13602.

\bibitem{kemelmacher2011exploring}
Kemelmacher-Shlizerman I, Shechtman E, Garg R, Seit, S~M (2011)
Exploring photobios. \textit{ACM Transactions on Graphics (TOG)} 30(4):61.

\bibitem{lim2013kinetics}
Lim B, Samper N, Lu H, Rushlow C, Jim{\'e}nez G, \textit{et~al.} (2013)
  Kinetics of gene derepression by erk signaling. \textit{Proc Natl Acad Sci USA} 110(25):10330--10335.

\end{thebibliography}

@@

\begin{figure}
\includegraphics{figS1}
\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green). The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right. The arrow indicates the position(s) where the cross-section of an embryo is imaged. (Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo. The dorsal (D) side is up and the ventral (V) side is down. For the first and third data sets (cellularization), images were collected at the focal plane $\sim 18\%$ from either the anterior or posterior pole of an embryo (arrows in the left image). During cellularization, it is known that the developmental dynamics are identical at these two regions \cite{lim2013kinetics}. In the second data set (gastrulation), images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image). }
\label{fig:ap_dv}
\end{figure}

\begin{figure}
\includegraphics{figS2}
\caption{{\it (A)} Images, stained for membranes (gray) and Dorsal (green) at different times during development. The membrane length is indicated by the red bars. {\it (B)} Membrane length as a function of time. The points corresponding to the images in {\it A} are indicated with arrows.}
\label{fig:membrane_compare}
\end{figure}

\begin{figure}
\includegraphics{figS3}
\caption{Images of mutant and wild type {\it Drosophila} embryos during cellularization, used for the analysis in \fig~5. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half are of mutant embryos.}
\end{figure}

\begin{figure}
\includegraphics{figS4}
\caption{Synthetic data set used to illustrate the data processing algorithms. {\it (A)} One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). {\it (B)} Intensity corresponding to the profile in {\it A}. {\it (C)} An ensemble of concentration profiles, each of the form described in {\it A}. Each row in the array corresponds to a single profile. {\it (D)} The profiles in {\it C}, now registered using angular synchronization. {\it (E)} The profiles in {\it D}, now temporally ordered using diffusion maps. {\it (F)} The profiles in {\it C}, registered and temporally ordered in a single step using vector diffusion maps.}
\label{fig:1d_demo}
%\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}{\it A}}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}{\it B}}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}{\it C}}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}{\it D}}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}{\it E}}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}{\it F}}
\end{figure}

\begin{figure*}
\includegraphics{figS5}
\caption{Illustration of the correspondence between three-dimensional rotations to translations and rotations in two dimensions. {\it (A)} The original image (shown in red), projected onto a portion of a sphere. {\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. {\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation.}
\label{fig:SO3_picture}
\end{figure*}

\begin{figure*}
\includegraphics{figS6}
\caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets.
%
We expect to see an approximately symmetric semicircle distribution centered at 0, which corresponds to noise, and positive eigenvalues separated from the semicircle which correspond to meaningful coordinates.
%
Each coordinate yields $d=3$ eigenvalues of (approximately) the same magnitude, and the top $d=3$ eigenvalues ($\lambda=1$) correspond to a trivial embedding coordinate.
%
(A) Histogram of eigenvalues from images during cellularization (\fig~3). As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics; then there is an appreciable gap before the third group of eigenvalues.
%
(B) Histogram of eigenvalues from images during gastrulation (\fig~4). We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are removed from this distribution (indicated by the red arrow). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first nontrivial VDM component which ``uncovers" the developmental dynamics.
%
(C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are removed from this distribution (indicated by the red arrows). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $6$ eigenvalues correspond to the first two VDM components which are now necessary to parametrize the developmental dynamics of the two genotypes.}
\label{fig:eval_spectra}
\end{figure*}

\begin{figure}
\includegraphics{figS7}
\caption{Images of embryos during cellularization (from \fig~3), registered using angular synchronization \cite{singer2011angular} and ordered by the first PCA projection coefficient. The recovered dynamics are qualitatively consistent with those recovered by vector diffusion maps. }
\label{fig:PCA_data1}
\end{figure}

\begin{figure*}
\includegraphics{figS8}

@@
